---
title: "Getting started with fullmetaphone"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fullmetaphone}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(fullmetaphone)
```

## Why phonetic codes

The same name is often spelled in many ways. Agarwal, Aggarwal and
Agrawal are one surname. Shree and Sri are one word. An exact match on a
name column misses all of these.

A phonetic code turns a name into a short string that stands for its
sound. Names that sound alike get the same code.

```{r}
metaphone(c("Agarwal", "Aggarwal", "Agrawal"))
```

## Two algorithms

`metaphone()` uses the original Metaphone algorithm of Lawrence Philips
(1990). It gives one code for each name.

`double_metaphone()` uses Philips' later Double Metaphone algorithm
(2000). It gives two codes. The primary code is the most likely
pronunciation. The secondary code allows for another common one.

```{r}
double_metaphone(c("Smith", "Schmidt", "Jose", "Michael"))
```

Smith and Schmidt share the code `XMT`, one as secondary and one as
primary. Double Metaphone treats two names as alike when any of their
codes match. `sounds_like()` does this check for you.

```{r}
sounds_like("Smith", "Schmidt")
sounds_like("Smith", "Schmidt", method = "metaphone")
```

## Why full-length codes

Most implementations keep only the first four characters of a code. For
short Western surnames that is enough. For long names it is not, and
Indian names are often long.

```{r}
names <- c("Venkatesh", "Venkataraman", "Venkatraman")
double_metaphone(names, max_length = 4)
double_metaphone(names)
```

With four characters, all three names look the same. With full codes,
Venkataraman and Venkatraman still match, as they should. Venkatesh is
now kept apart.

The difference adds up quickly. In one test set of 5000 synthetic Indian
names, four-character codes gave about 2700 distinct primary codes. Full
codes gave about 4950. So four-character codes would have flagged many
pairs of different people as possible duplicates.

Use `max_length = 4` whenever you need the traditional short codes, for
example to compare with results from other software.

## Names with several words

By default the whole name is coded as one string, as in the original
algorithm. Set `by_word = TRUE` to code each word on its own. The word
codes are joined with spaces.

```{r}
double_metaphone("Ramesh Kumar Sharma")
double_metaphone("Ramesh Kumar Sharma", by_word = TRUE)
```

Word-by-word codes help when the parts of a name can come in a different
order. You can sort the word codes before comparing them.

```{r}
codes <- metaphone(c("Ramesh Kumar", "Kumar Ramesh"), by_word = TRUE)
codes
sorted <- vapply(strsplit(codes, " "),
                 function(w) paste(sort(w), collapse = " "),
                 character(1))
sorted[1] == sorted[2]
```

## A small de-duplication example

Here is a list of payees with some planted near-duplicates.

```{r}
payees <- data.frame(
  id = 1:10,
  name = c("Shree Ganesh Traders", "Shri Ganesh Traders",
           "Venkataraman Enterprises", "Venkatraman Enterprises",
           "Venkatesh Enterprises",
           "Agarwal and Sons", "Agrawal and Sons",
           "Mohan Lal", "Mohan Laal", "Sohan Lal")
)
payees <- cbind(payees, double_metaphone(payees$name, by_word = TRUE))
payees
```

Group the rows by their primary code and keep the groups with more than
one row.

```{r}
groups <- split(payees$id, payees$primary)
groups[lengths(groups) > 1]
```

Four pairs are flagged. Shree and Shri, Venkataraman and Venkatraman,
Agarwal and Agrawal, and Lal and Laal. Venkatesh Enterprises and Sohan
Lal are rightly left alone. A real review would also compare the
secondary codes, and would then check each flagged pair by hand.

## What phonetic codes miss

Phonetic codes are not magic. They follow English spelling rules, and
some variations defeat them.

```{r}
double_metaphone(c("Shree", "Sri"))$primary
double_metaphone(c("Agarwal and Sons", "Agarwal & Sons"))$primary
double_metaphone(c("Trader", "Traders"))$primary
```

"Sri" starts with S and "Shree" starts with SH, so they get different
codes. An ampersand is dropped while "and" is coded. A plural adds an S.
For such cases, standardise the text first. Replace "&" with "and",
"Sri" with "Shree", remove words like "Traders" or "M/s", and only then
compute the codes.

## Preparing the text

Both functions change letters to capitals. They replace common accented
Latin letters with plain ones, drop apostrophes, and treat other
punctuation and digits as spaces. Missing values stay missing.

```{r}
double_metaphone(c("O'Brien", "D'Souza", NA, "123"))
```

The algorithms work on Latin letters only. Names in Devanagari or other
scripts should be transliterated first.

## Credits

Lawrence Philips designed both algorithms. The Double Metaphone rules
here follow the C implementation by Maurice Aubrey, with fixes by Kevin
Atkinson. The 'PGRdup' package by J. Aravind and co-authors uses the same
C code and first brought Double Metaphone to R. Its
`DoubleMetaphone()` function returns four-character codes. The Metaphone
rules and the reference test data come from Apache Commons Codec.

## References

Philips, L. (1990). Hanging on the metaphone. *Computer Language*,
7(12), 39-44.

Philips, L. (2000). The double metaphone search algorithm. *C/C++ Users
Journal*, 18(6), 38-43.
